---
title: "problem-set6"
author: "jubin"
date: "August 24, 2017"
output: html_document
---

### Quiz 1: price vs. x
```{r}
library(ggplot2)
data("diamonds")
ggplot(aes(x=price,y=x), data=diamonds)+
  geom_point(alpa=0.1)
```

### Quiz 2: Findings - price vs. x
```{r}
'Relationship of X vs. Price is Exponential and there are several outliers.'
```

### Quiz 3: Correlations
```{r}
cor.test(diamonds$price,diamonds$x)
cor.test(diamonds$price,diamonds$y)
cor.test(diamonds$price,diamonds$z)
```

### Quiz 4: price vs. depth
```{r}
ggplot(aes(x=price,y=depth), data=diamonds)+
  geom_point(alpa=0.1)
```

### Quiz 5: Adjustments - price vs. depth
```{r}
ggplot(data = diamonds, aes(x = depth, y = price)) + 
  geom_point(alpha=0.01)+
  scale_x_continuous(limits=c(50, 70), breaks=seq(50, 70, 2))
```

### Quiz 6: Typical Depth Range
```{r}
'Lower Limit: 58
Upper Limit: 64'
```

### Quiz 7: Correlation - price and depth
```{r}

```

### Quiz 8: price vs. carat
```{r}
```

### Quiz 9: price vs. volume
```{r}
```

### Quiz 10: Findings - price vs. volume
```{r}
```

### Quiz 11: Correlations on Subsets
```{r}
```

### Quiz 12: Adjustments - price vs. volume
```{r}
```

### Quiz 13: Mean pice by clarity
```{r}
```

### Quiz 14: Bar Charts of Mean Price
```{r}
```

### Quiz 15: Trends in Mean Price
```{r}
```

### Quiz 16: Gapminder Revisited
```{r}
```